---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<!-- badges: start -->
[![License: GPL v3](https://img.shields.io/badge/License-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
[![R build status](https://github.com/PolMine/bignlp/workflows/R-CMD-check/badge.svg)](https://github.com/PolMine/bignlp/actions)
[![codecov](https://codecov.io/gh/PolMine/bignlp/branch/master/graph/badge.svg)](https://codecov.io/gh/PolMine/bignlp/branch/master)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
<!-- badges: end -->
  


## The 'bignlp'-package: Objectives

R users who want to add a linguistic annotation to a corpus are already offered
a few available packages for Natural Language Processing (NLP):

- [openNLP](https://CRAN.R-project.org/package=openNLP)
- [cleanNLP](https://CRAN.R-project.org/package=cleanNLP)
- [coreNLP](https://CRAN.R-project.org/package=coreNLP)
- [spacyr](https://CRAN.R-project.org/package=spacyr)
- [udpipe](https://CRAN.R-project.org/package=udpipe)

These packages are not "pure R" NLP-tools. They offer interfaces to standard NLP
tools implemented in other programming languages. The
[cleanNLP](https://CRAN.R-project.org/package=cleanNLP) R package offers to
combine these external tools in one coherent framework. So why yet another NLP R
package?

The listed packages are not good to deal with large volumes of text as a matter
of performance and memory efficiency. The bignlp R package offers fast, robust
and convenient functionality for annotating large-scale corpora, including XML
data formats.  It uses [Stanford
CoreNLP](https://stanfordnlp.github.io/CoreNLP/) as standard, high-quality tool
in parallel mode. To be parsimonious with the memory available, it implements
line-by-line processing, so that annotated data is not be kept in memory. The
bignlp package may be your package of choice you if your data is substantially
big.



## Installation and dependencies

The bignlp package itself is a pure R package. It does not include compiled
code, and the installation of the package itself is unlikely to cause problems.
At this stage, bignlp is a GitHub-only package. Using an installation mechanism
of the [remotes]() package exported by [devtools]() you can install bignlp as
follows.

```{r}
devtools::install_github("PolMine/bignlp")
```

However bignlp depends on:

- The  Java library [Stanford CoreNLP](https://stanfordnlp.github.io/CoreNLP/)
- The availability of a Java Virtual Machine. Java 1.8 is recommended.
- [rJava](https://CRAN.R-project.org/package=rJava)


### rJava and Java

Install the rJava R package:

```{r, eval = FALSE}
install.packages("rJava")
```

Stanford CoreNLP officially requires Oracle Java v1.8+, see [the CoreNLP
Website](https://stanfordnlp.github.io/CoreNLP/). Using CoreNLP in combination
with OpenJDK, a Java flavour that is available on many systems, is not
officially supported. Yet we are not aware of errors resulting from using the
"wrong" Java. If you see errors, you might want to ensure that you have the
"correct"" Java (Java Development Kit / JDK) installed.

Oracle Java can be installed from [the Oracle
website](https://www.oracle.com/technetwork/java/javase/downloads/jdk8-downloads-2133151.html).
Note that Oracle licenses have become more restrictive, and registration may be
necessary. This will not be a problem for academic/non-commercial contexts.

Once downloading and installing Oracle Java 1.8+ is completed, making this Java
version the one that is actually used may be tricky. For macOS, we found the
jEnv tool to be useful, see [this
link](https://medium.com/@danielnenkov/multiple-jdk-versions-on-mac-os-x-with-jenv-5ea5522ddc9b).

To get R use Java 1.8, you may finalle have to call R CMD javareconf from the command line.

```{sh, eval = FALSE}
R CMD javareconf
```

To check within R that everything works, run this snippet from within R.

```{r, eval = FALSE}
library(rJava)
.jinit()
.jcall("java/lang/System", "S", "getProperty", "java.runtime.version")
```

The output you see (something like "1.8.0_211-b12") on your system depend on the
Java version you have installed.


### Installing CoreNLP

CoreNLP is a Java package that needs to be present on your machine. 

A conventional place for installing a tool such as CoreNLP on Linux and macOS is
the /opt dir. So from a terminal, create a directory for CoreNLP, go into it,
download the zipped jar files, unzip it, and remove the zip file. Note that sudo
rights may be necessary to have permissions to write into the /opt directory.

```{sh, eval = TRUE}
CORENLP_LATEST=`Rscript -e 'bignlp::corenlp_latest()'|grep -oE '\d+\.\d+\.\d+'`
```

```{sh download_corenlp, eval = FALSE}
mkdir /opt/stanford-corenlp
cd /opt/stanford-corenlp
wget http://nlp.stanford.edu/software/stanford-corenlp-$CORENLP_LATEST.zip
unzip stanford-corenlp-$CORENLP_LATEST.zip
rm stanford-corenlp-$CORENLP_LATEST.zip
```

You may want to to install a language model for English.

```{sh download_english_model, eval = FALSE}
cd /opt/stanford-corenlp/stanford-corenlp-$CORENLP_LATEST
wget http://nlp.stanford.edu/software/stanford-corenlp-$CORENLP_LATEST-models-english.jar
```

We illustrate getting models for a specific language for German. We go into the
CoreNLP directory and download the model to this place.

```{sh download_german_model, eval = FALSE}
cd stanford-corenlp-$CORENLP_LATEST
wget http://nlp.stanford.edu/software/stanford-corenlp-$CORENLP_LATEST-models-german.jar
```

The jar file with the model include a default properties file for processing
German data (StanfordCoreNLP-german.properties). You can see this by displaying
the content of the jar as follows.

```{sh inspect_jar, eval = FALSE}
jar tf stanford-corenlp-$CORENLP_LATEST-models-german.jar | grep "properties" # see content of jar
```

The inclusion of this properties file in the jar may become a problem. If you want to configure the parser yourself, you may encounter the issue that this properties file included in the jar will override any other properties you may want to use. 

The solution we found to work is to (a) extract the properties file from the jar and (b) remove it from the jar.

```{sh modify_model_jar, eval = FALSE}
# extract StanfordCoreNLP-german.properties from jar
unzip stanford-corenlp-$CORENLP_LATEST-models-german.jar StanfordCoreNLP-german.properties
# remove StanfordCoreNLP-german.properties from jar
zip -d stanford-corenlp-$CORENLP_LATEST-models-german.jar StanfordCoreNLP-german.properties
```

Note that the bignlp package already includes a properties file that has been
edited for annotating large amounts of data quickly. You will find it as
follows:

```{r pkg_properties_file, eval = FALSE}
options(bignlp.properties_file = corenlp_get_properties_file(lang = "de", fast = TRUE))
```


## Core Functionality

```{r}
library(bignlp)
library(data.table)

reuters_file <- system.file(package = "bignlp", "extdata", "txt", "reuters.txt")
reuters_txt <- readLines(reuters_file)
reuters_tab <- data.table(
  doc_id = 1L:length(reuters_txt), 
  text = reuters_txt
)

props <- corenlp_get_properties_file(lang = "en", fast = "TRUE")

y <- corenlp_annotate(
  x = reuters_tab,
  pipe = props,
  corenlp_dir = corenlp_get_jar_dir(),
  progress = FALSE
)
```

## Known issues

### rJava

An issue we have seen on macOS in combination with RStudio is that the rJava
dynamic library called from the RStudio session does not link correctly against
the dynamic library of the Java virtual machine. We use the follwing snippet to
induce the correct link. (Note that paths may have to be updated to reflect the
correct R and Java version!)

```{sh, eval = FALSE}
install_name_tool -change "@rpath/libjvm.dylib" \
  "/Library/Java/JavaVirtualMachines/jdk1.8.0_211.jdk/Contents/Home/jre/lib/server/libjvm.dylib" \
  /Library/Frameworks/R.framework/Versions/3.5/Resources/library/rJava/libs/rJava.so
```


### Issues with RStudio

We have seen problems with the parallel annotation with from an RStudio session.
Running bignlp from a simple terminal window is a good solution. For a large
annotation project, omitting an IDE may be a good idea anyway.
